---
title: "SVM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SVM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(ClassifiR)
library(ggplot2)
library(readxl)
```

## Introduction

This vignette demonstrates how to use the `my_svm()` function from the `ClassifiR` package.

Unlike black-box implementations in common machine learning packages, `my_svm()` is built from scratch using stochastic gradient descent (SGD) to train a linear Support Vector Machine (SVM) classifier. This makes it highly suitable for learning and teaching purposes, giving insight into the mechanics of margin-based classification.

Key features of `my_svm()` include:

1. Outputs both predicted labels and probabilities (via sigmoid of margin);

2. Visibility into raw margin values and model weights.

In addition to basic classification, it supports automatic hyperparameter tuning for both `Learning rate (learning_rate)` and `Regularization strength (lambda)`.

## Load and Prepare Data

We'll use the preprocessed weather dataset provided in the package. It is already cleaned, normalized, and converted to a binary classification format. The last column in each dataset is a binary label (e.g., rain or no rain), and all other columns are numeric features normalized to [0, 1].

```{r}
data("weatherAUS_processed")
train_test_list = splitDataset_(weatherAUS_processed)
train_data = train_test_list$train
test_data = train_test_list$test
```

## Train the SVM

Now we train the linear SVM classifier using `my_svm()`. You can experiment with `learning_rate`, `lambda` (regularization strength), and `n_iter` (number of passes through the data).

You can use the default parameters to train a simple linear SVM without tuning.
```{r}
model <- my_svm(train_data, test_data, learning_rate = 0.01, lambda = 0.01, n_iter = 50)
```

## Train with Parameter Tuning

Now, we train the same model with automatic tuning enabled. It will internally search for the best combination of learning rate and lambda using a validation set.

`learning_rates` is a numeric vector of learning rates to try (e.g., c(0.001, 0.01)). Default is 0.01.

`lambda_values` is a numeric vector of lambda (regularization) values to try. Default is c(0.0001, 0.001, 0.01, 0.1).

```{r}
model_tuned <- my_svm(train_data, test_data,
                      learning_rates = c(0.001, 0.005, 0.01),
                      lambda_values = c(0.001, 0.01, 0.1),
                      n_iter = 100,
                      tune = TRUE)
```

## Predictions result
The returned object contains:

`weights`: learned coefficients (including intercept),

`prediction_train`, `prediction`: predicted labels (0 or 1),

`probability_train`, `probability`: estimated probabilities from the sigmoid of margins,

`train_margins`, `margins`: raw margin scores used in classification,

`predict_fun`: the prediction function itself (for new data).

```{r}
# View prediction results
head(model$prediction)
head(model$probability)
```

We can evaluate test accuracy and inspect predicted values:

```{r}
accuracy_svm <- mean(model$prediction == test_data[[ncol(test_data)]])
accuracy_svm
```

## Visualization
Margins are the core of SVM theory — they indicate how far a sample lies from the decision boundary. We can plot the predicted probabilities (derived from margins via the sigmoid function) to see how confidently the model makes its decisions.

```{r}
# Plot margins and probabilities
df_svm <- data.frame(
  Actual = factor(test_data[[ncol(test_data)]]),
  Predicted = model$prediction,
  Probability = model$probability
)

ggplot(df_svm, aes(x = Probability, fill = Actual)) +
  geom_histogram(bins = 15, alpha = 0.6, position = "identity") +
  labs(title = "SVM: Predicted Probabilities (based on margins) vs Actual Labels",
       x = "Predicted Probability",
       fill = "Actual Label")
```

## Conclusion

The `my_svm()` function in `ClassifiR` is an educational tool for mastering the principles behind Support Vector Machines. Its transparent architecture enables:

1. Step-by-step control over model fitting,

2. Inspection of margins and weights,

3. Customization and parameter tuning for performance analysis.

Use this function to build intuition about margin-based learning and the trade-offs involved in model complexity and regularization.
